# Existing Vegetation Types {#evt}

While looking at the historical ecosystems gives context, we also need to get a picture of which ecosystems are on the landscape today.  



## General methods

We use the same general methods as we did on the "Historical Ecosystems" page, using a Pivot Table.  We rely on [LANDFIRE's Existing Vegetation Type (EVT)](https://www.landfire.gov/evt.php) data for this assessment.


## Getting at amounts of current ecosystems

**Start by opening the "current" tab in the Excel workbook.**

1. In the Pivot Table Fields pane, select "EVT_NAME" then "ACRES".  Make sure that EVT_NAME is in the "Rows" pane, and that "Sum of ACRES" is in the "Values" pane.  
2. Right click in the top cell of the "Sum of ACRES" column (not the column header) in the Pivot Table, then "Sort Largest to Smallest".
3. In our example we have some BpSs that have low ACRES values.  We also have categories that are not meaningful, such as "Barren-Rock/Sand/Clay".  We can do a little formatting/cleaning before making a chart:
    * To remove EVTs from the table you will click the drop-down menu to the right of "BPS_NAME" in the Pivot Table Fields pane.  You can uncheck BpSs as appropriate.
    * It is also possible to filter by right clicking on the top value in the list of EVTs within the Pivot Table, then selecting Filter > Top 10....  Once in that menu you can refine the filtering.
    * Additionally we like to add commas and remove decimal places.  To do this right click the row that contains "Sum of ACRES" in the spreadsheet (likely row "B"), select "Format Cells" > Number > set Decimal places to "0" then check the "Use 1000 Separator box.
4. To get percentages, drag "ACRES" from the top Pivot Table Field pane to the "Values" pane.  This will add a second "ACRES" column to the table.  Click the drop down in the second instance of "ACRES" (reads "SUM of ACRES2" in our example), then Value Field Settings.  In this menu select the "Show Values As" tab, click the "Show Values As" drop down then select "% of Grand Total% to get percentages of each BpS (make sure that "BPS_NAME" is selected as the "Base field").  
5.  To get a "running total" of percentages you will add a third instance of "ACRES" to the "Values" pane, then Value Field Settings.  In this menu select the "Show Values As" tab, click the "Show Values As" drop down then select "% Running Total In" to get running totals of  percentages of each BpS (make sure that "BPS_NAME" is selected as the "Base field"). 
6. Save!




**Example output**

Below is essentially the output from our Pivot Table work with a couple changes:

* Names have been shortened for ease of reading
* Numbers have been rounded

```{r evtDT, echo=FALSE, message=FALSE, warning=FALSE}
library(DT)
library(readr)
evt <- read_csv("evt.csv")


datatable(head(evt), 
          class = 'cell-border stripe',
          options = list(scrollX  = TRUE, pageLength = 3))%>%
  formatRound(c("ACRES", "PERCENT", "RUNNING PERCENT"), 0)
```

<br>
<br>

We see that first there are many EVT's.  This partially due to the addition of Developed and other "modern" categories such as "Recently Logged-Tree Cover". Second we see that of the "natural" types their is a close match with the historical.  We can visually confirm this and other patterns with a quick chart made in R (similar charts available in Excel by highlighting the data, clicking Insert then selecting the chart type in the "Charts" tab):

<br>

```{r evtChart, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(scales)
library(dplyr)

evt %>%
  top_n(10, ACRES) %>%
ggplot() +
  aes(x = reorder(EVT, ACRES), y = ACRES) + 
  geom_bar(stat = "identity", fill = "grey31") +
  coord_flip() +
  labs(x = "", 
       y = "Acres", 
       title = "Current Ecosystems of the Michigamme Highlands",
       caption = "From LANDFIRE Existing Vegetation Type data. Names shortened for clarity",
       colour = "black") +
  scale_y_continuous(labels = comma) +
theme(plot.title = element_text(size = 15), axis.title = element_text(size = 13), panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
    theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot")
```


