--- 
title: "Using LANDFIRE Products to explore historical and current ecosystems"
author: "Draft by The Nature Conservancy's LANDFIRE team"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book



---

# This document...

is a technical guide originally created for The Nature Conservancy's Global Science Gathering, working session titled **"Billions (and billons) of pixels coupled with hundreds of ecosystem models: Making LANDFIRE products work for your landscape"** delivered by Kori Blankenship and Randy Swaty.  It was written in R using [R-Studio](https://rstudio.com/), the ["bookdown" package](https://www.bookdown.org/) and is hosted on [GitHub(links to this repository)](https://github.com/rswaty/landfireFSC).  

You are a guinea pig, the subject of an experiment.  We hope you do not feel duped.  We'd sincerely appreciate your feedback on the session, any ideas or edits you have for this document and questions you may have about what we do.  Please e-mail use at [landfire@tnc.org](landfire@tnc.org).  

The goals are to:

* get acres of ecosystems past and present
* explore ecosystem conversion
* compare amounts of succession classes past and present


In the United States, including the insular areas, [LANDFIRE](https://www.landfire.gov/) provides the datasets and ecological model results to get at these challenges and more.  Here we walk you through some of the technical steps needed to start your analysis.  We will do our work in a model landscape, the Michigamme Highlands in the Upper Peninsula of Michigan (highlighted in green in map below).

To help you with the concepts we work through pre-processed LANDFIRE data for an example landscape.  To prepare the datasets for your landscape see the Appendix.


**ZOOM, pan, explore the Michigamme Highlands area**
```{r loadshp, message=FALSE, warning=FALSE, include=FALSE}
library(raster)
library(leaflet)
library(sf)

michi <- st_read("michishp.shp")
michi <- st_transform(michi, CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

```


```{r leafletMichi, echo=FALSE, message=FALSE, warning=FALSE}

leaflet(data = michi) %>% 
  addTiles() %>% 
  addPolygons(fill = FALSE, stroke = TRUE, color = "#193b22")

```


## Download this!
To work through this guide you will need to download this [zipped folder](https://github.com/rswaty/landfireFSC/blob/main/toDownload.zip) by clicking the hyperlinked text, then the "Download" button that will be on the right side of the screen.  This folder contains:

* An Excel file
* One ecosystem description
* An ArcMap map package file in case you want to explore in ArcMap.










